
---
title: "Final Project R"
author: "Alessandra Bielli"
date: "2023-12-07"
output:
  html_document:
    code_folding: hide
    self_contained: true
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (dplyr)
library(tidyverse)
library(readr)
library(tibble)
library(stringr)
library(ggplot2)
library(tidyr)
```
 
## Step 1: Load the dataset into RStudio and convert to a tibble
```{r}
# Load the CSV file into a data frame
setwd("/Users/Allie/Desktop/")
data <- read.csv("Austin_Animal_Center_Outcomes.csv")

# Convert to a tibble (optional)
animal_tibble <- as_tibble(data)

# Remove columns Date/time and Outcome.Subtype
animal_tibble <- animal_tibble %>%
  select(-DateTime, -Outcome.Subtype)

# Display the first few rows of the updated tibble
head(animal_tibble)
```
## Step 2: Tidy the data for easier analysis using  tidyr and dplyr
```{r}
# Update column names in animal_tibble
animal_tibble <- animal_tibble %>%
  rename(
    Animal_ID = Animal.ID,  
    Date_of_Birth = Date.of.Birth,
    Outcome_Type = Outcome.Type,
    Animal_Type = Animal.Type,
    Sex_At_Outcome = Sex.upon.Outcome,
    Year_At_Outcome = Age.upon.Outcome,
    Date_of_Outcome = MonthYear
  )

# Remove rows with Animal_Type 'Other'
cleaned_animal_tibble <- animal_tibble %>%
  filter(Animal_Type != "Other")

# Fill remaining blank spaces with "Unknown"
filled_cleaned_tibble <- cleaned_animal_tibble %>%
  mutate_all(~ifelse(. == "" | is.na(.), "Unknown", .))

# Threshold for unknown values
threshold <- 1  # Define the threshold here

# Remove rows with more than 'threshold' unknown values
cleaned_animal_tibble_filtered <- cleaned_animal_tibble %>%
  filter(rowSums(. == "" | is.na(.)) <= threshold)

# Sort by 'MonthYear' 
filled_cleaned_tibble <- filled_cleaned_tibble %>%
  arrange(Date_of_Outcome, Name)

# Convert Date_of_Birth column to Date format
filled_cleaned_tibble <- filled_cleaned_tibble %>%
  mutate(Date_of_Birth = as.Date(Date_of_Birth, format = "%m/%d/%y"))

# Check unique values in Year_At_Outcome column
unique(filled_cleaned_tibble$Year_At_Outcome)

# Filter rows with negative ages
filtered_data <- filled_cleaned_tibble %>%
  filter(Year_At_Outcome >= 0)

# Clean Year_At_Outcome column for numeric analysis
filled_cleaned_tibble <- filled_cleaned_tibble %>%
  mutate(Year_At_Outcome = as.numeric(gsub(" years", "", Year_At_Outcome)))

# Check unique values in Year_At_Outcome column that couldn't be converted
non_convertible_values <- unique(filled_cleaned_tibble[is.na(as.numeric(gsub(" years", "", filled_cleaned_tibble$Year_At_Outcome))), "Year_At_Outcome"])
non_convertible_values

# Now, Year_At_Outcome can be used for numerical analysis

# Remove rows with NA in Year_At_Outcome column
filled_cleaned_tibble <- filled_cleaned_tibble %>%
  filter(!is.na(Year_At_Outcome))

# Convert Date_of_Outcome column to desired format
library(zoo)
filled_cleaned_tibble <- filled_cleaned_tibble %>%
  mutate( Date_of_Outcome = as.yearmon(Date_of_Outcome, "%b-%y")) %>%
  mutate( Date_of_Outcome = format(as.Date(Date_of_Outcome), "%Y-%m"))

# View the first few rows after sorting
head(filled_cleaned_tibble)

# Check the number of rows in the filled_cleaned_tibble dataset
n_rows <- nrow(filled_cleaned_tibble)
print(n_rows)
```

## Step 3: Display a bar chart showing count for each animal type since 2013
```{r}
library(ggplot2)

# Create a bar plot for count of each animal type
ggplot(filled_cleaned_tibble, aes(x = Animal_Type)) +
  geom_bar(fill = "skyblue", color = "black") +
  geom_text(
    aes(label = after_stat(count)),  # Use after_stat(count) instead of stat(count)
    stat = "count", 
    vjust = -0.5
  ) +
  labs(title = "Count of Each Animal Type", x = "Animal Type", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability if needed
```

## Step 4: Create a histogram displaying the outcomes for birds and livestock
```{r}
library(dplyr)
library(ggplot2)

# Filter data for birds and livestock
birds_livestock_data <- filled_cleaned_tibble %>%
  filter(Animal_Type %in% c("Bird", "Livestock"))

# Calculate total counts for birds and livestock
total_counts <- birds_livestock_data %>%
  count(Animal_Type)

# Calculate percentages for each outcome of birds and livestock
outcome_percentages <- birds_livestock_data %>%
  group_by(Animal_Type, Outcome_Type) %>%
  summarise(count = n()) %>%
  left_join(total_counts, by = "Animal_Type") %>%
  mutate(percentage = (count / n) * 100) %>%
  filter(percentage != 0)  # Exclude outcomes with 0%

# Create a bar plot for outcome percentages of birds and livestock
ggplot(outcome_percentages, aes(x = Outcome_Type, y = percentage, fill = Animal_Type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  labs(title = "Outcome Percentages of Birds and Livestock",
       x = "Outcome Type", y = "Percentage") +
  theme_minimal()
```

## Step 5: Create a histogram displaying the outcomes for cats and dogs  
```{r}
library(dplyr)
library(ggplot2)

# Filter data for cats and dogs
cats_dogs_data <- filled_cleaned_tibble %>%
  filter(Animal_Type %in% c("Cat", "Dog"))

# Calculate outcome percentages for cats and dogs
outcome_percentages <- cats_dogs_data %>%
  count(Animal_Type, Outcome_Type) %>%
  group_by(Animal_Type) %>%
  mutate(percentage = n / sum(n) * 100)

# Create a bar plot for outcome percentages of cats and dogs with filtered text labels
ggplot(outcome_percentages, aes(x = Outcome_Type, y = percentage, fill = Animal_Type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(data = outcome_percentages %>%
              filter(percentage >= 0.01),  # Filter labels below 0.01%
            aes(label = paste0(round(percentage), "%")),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  labs(title = "Outcome Percentages for Cats and Dogs",
       x = "Outcome Type", y = "Percentage") +
  theme_minimal()
```

## Step 6: Create a violin graph displaying the age distribution of cats and dogs
```{r}
 library(ggplot2)

# Filter the data for relevant columns and animal types
age_data <- filled_cleaned_tibble %>%
  filter(Animal_Type %in% c("Cat", "Dog")) %>%
  select(Animal_Type, Year_At_Outcome)

# Create a violin plot for age distribution by animal type
ggplot(age_data, aes(x = Animal_Type, y = Year_At_Outcome, fill = Animal_Type)) +
  geom_violin() +
  labs(title = "Age Distribution by Animal Type",
       x = "Animal Type", y = "Age at Outcome") +
  theme_minimal()
```

## Display a histogram showing age group percentages among cats and dogs 
```{r}
library(dplyr)
library(ggplot2)

# Filter data for cats and dogs
cats_dogs_data <- filled_cleaned_tibble %>%
  filter(Animal_Type %in% c("Cat", "Dog"))

# Calculate total counts for cats and dogs
total_counts <- cats_dogs_data %>%
  count(Animal_Type)

# Define age groups
cats_dogs_data <- cats_dogs_data %>%
  mutate(Age_Group = cut(Year_At_Outcome, breaks = c(1, 3, 5, 10, 15, 20, Inf),
                         labels = c("1-3", "3-5", "5-10", "10-15", "15-20", "20+"),
                         include.lowest = TRUE))

# Calculate percentages for each age group of cats and dogs
age_group_percentages <- cats_dogs_data %>%
  group_by(Animal_Type, Age_Group) %>%
  summarise(count = n()) %>%
  left_join(total_counts, by = "Animal_Type") %>%
  mutate(percentage = (count / n) * 100) %>%
  filter(percentage != 0)  # Exclude age groups with 0%

# Create a bar plot for age group percentages of cats and dogs
ggplot(age_group_percentages, aes(x = Age_Group, y = percentage, fill = Animal_Type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  labs(title = "Age Group Percentages of Cats and Dogs",
       x = "Age Group", y = "Percentage") +
  theme_minimal()
```

## Display percentage distribution of age groups for cats and dogs within each outcome 
```{r}
library(dplyr)
library(ggplot2)

# Filter data for cats and dogs
cats_dogs_data <- filled_cleaned_tibble %>%
  filter(Animal_Type %in% c("Cat", "Dog"))

# Define age groups
cats_dogs_data <- cats_dogs_data %>%
  mutate(Age_Group = cut(Year_At_Outcome, breaks = c(1, 3, 5, 10, 15, 20, Inf),
                         labels = c("1-3", "3-5", "5-10", "10-15", "15-20", "20+"),
                         include.lowest = TRUE))

# Calculate percentages for each age group of cats and dogs within each outcome
age_group_percentages <- cats_dogs_data %>%
  count(Outcome_Type, Animal_Type, Age_Group) %>%
  group_by(Outcome_Type, Animal_Type) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  filter(!is.na(Age_Group) & percentage != 0)  # Exclude rows with NA in Age_Group and 0% percentage

# Create a bar plot for age group percentages of cats and dogs within each outcome
ggplot(age_group_percentages, aes(x = Age_Group, y = percentage, fill = Animal_Type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  facet_wrap(~ Outcome_Type, scales = "free_x", strip.position = "bottom") +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_dodge(width = 0.8), vjust = -0.8, size = 2) +
  labs(title = "Percentage Distribution of Age Groups for Cats and Dogs within Each Outcome",
       x = "Age Group", y = "Percentage") +
  theme_minimal()
```


## Display a line graph and scatter plot showing outcome trends for cats and dogs over time 
```{r}
library(dplyr)
library(ggplot2)
library(zoo)

# Filter data for cats and dogs
cats_dogs_data <- filled_cleaned_tibble %>%
  filter(Animal_Type %in% c("Cat", "Dog"))

# Extract the year from the Date_of_Outcome column
cats_dogs_data <- cats_dogs_data %>%
  mutate(Year = as.yearmon(Date_of_Outcome))

# Calculate outcome percentages for cats and dogs over years
outcome_percentages_over_time <- cats_dogs_data %>%
  count(Animal_Type, Year, Outcome_Type) %>%
  group_by(Animal_Type, Year) %>%
  mutate(percentage = n / sum(n) * 100)

# Create a time series plot for outcome percentages of cats and dogs over years
ggplot(outcome_percentages_over_time, aes(x = Year, y = percentage, color = Outcome_Type)) +
  geom_line() +
  facet_wrap(~Animal_Type, scales = "free_y") +
  labs(title = "Outcome Trends of Cats and Dogs Over Time",
       x = "Year", y = "Percentage") +
  theme_minimal()

# Assuming you've previously filtered and manipulated the data

# Create a scatter plot for outcome percentages of cats and dogs over years
ggplot(outcome_percentages_over_time, aes(x = Year, y = percentage, color = Outcome_Type)) +
  geom_point() +
  facet_wrap(~Animal_Type, scales = "free_y") +
  labs(title = "Outcome Trends of Cats and Dogs Over Time",
       x = "Year", y = "Percentage") +
  theme_minimal()

```
